Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-31 results.tex
edits,figures and variables

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:wwpe}.  The objective of these
experiments was to explore how prediction accuracy is related to WPE,
and how that relationship depends on the generating process and the
prediction method.  Working from the first 90\% of each signal, we
generated a prediction of the last 10\% using the \naive, ARIMA, and
LMA prediction methods, as described in Section~\ref{sec:accuracy},
then calculated the MASE value of those predictions.  We also
calculated the WPE of each time series using a wordlength of six, as
described in Section~\ref{sec:meaComplex}.  In order to assess the
run-to-run variability of these results, we repeated all of these
calculations on 15 separate trials: i.e., 15 different runs of each
program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of
these 45 predictions for each system: i.e., the lowest error over all
15 trials and all three methods for each of the eight programs.  The
WPE is plotted against the corresponding MASE value here in order to
bring out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.  [[Ryan, is that correct?]]}
  \label{fig:wpe_vs_mase_best}
\end{figure}

>>>>

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective of these
experiments was to explore how prediction accuracy is related to WPE,
and how that relationship depends on the generating process and the
prediction method.  Working from the first 90\% of each signal, we
generated a prediction of the last 10\% using the \naive, ARIMA, and
LMA prediction methods, as described in Section~\ref{sec:accuracy},
then calculated the MASE value of those predictions.  We also
calculated the WPE of each time series using a wordlength of six, as
described in Section~\ref{sec:meaComplex}.  In order to assess the
run-to-run variability of these results, we repeated all of these
calculations on 15 separate trials: i.e., 15 different runs of each
program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of
these 45 predictions for each system: i.e., the lowest error over all
15 trials and all three methods for each of the eight programs.  The
WPE is plotted against the corresponding MASE value here in order to
bring out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.  [[Ryan, is that correct?]]}
  \label{fig:wpe_vs_mase_best}
\end{figure}

### 2014-03-31 wpe.tex
edits,figures and variables

Utilizing Shannon entropy rate as a measure of temporal complexity is
not a new idea \cite{Shannon1951, mantegna1994linguistic}, but
previous approaches required categorical data: $x_t \in \mathcal{S}$
for some finite or countably infinite \emph{alphabet} $\mathcal{S}$.
Data taken from real-world systems is, however,
effectively\footnote{Measurements from finite-precision sensors are
  discrete, but data from modern high-resolution sensors are, for the
  purposes of entropy calculations, effectively continuous.}
real-valued.  To analyze real-valued data using a method that requires
categorical values, one must discretize the data---typically by
binning.  Unfortunately, this is rarely a good solution, as the
binning of the values introduces spurious dynamics~\cite{bollt2001}.
The field of symbolic dynamics offers discretization methods that do
not disturb the intrinsic behavior.  These methods are, however,
fragile in the face of noise; worse yet, they require knowledge of the
underlying system.  This is not useful in our context, where we want
to measure the entropy to quantify predictability.

@@

  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_\ell$ as all $\ell!$ permutations $\pi$ of order $\ell$. For each $\pi \in \mathcal{S}_\ell$ we determine the relative frequency of that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$:
  \begin{align*}
    P(\pi) = \frac{\left|\{t|t \leq T-\ell,\phi(x_{t+1},\dots,x_{t+\ell}) = \pi\}\right|}{T-\ell+1}
  \end{align*}
  where $P(\pi)$ quantifies the probability of an ordinal and
  $|\cdot|$ is set cardinality. The \emph{permutation entropy} of
  order $\ell \ge 2$ is defined as
  \begin{align*}
    H(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P(\pi) \log_2 P(\pi)
  \end{align*}

@@

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  The intent behind the weighting
is to correct for observational noise that is larger than the trends
in the data, but smaller than the larger scale features.  Consider,
for example, a signal that switches between two fixed points and
contains some additive noise. The PE is dominated by the noise about 
the fixed points driving it to be $\approx 1$, in some sense hiding
the fact that the signal is actually quite structed. The weighted 
permutation entropy on the other hand would be dominated by the 
switching rather than by the stochastic fluctuations. To accomplish 
this, the \emph{weight} of a permutation is taken into account:
\begin{align*}
  w(x_{t+1}^\ell) = \frac{1}{\ell} \sum_{i = t}^{t+\ell}
                      \left( x_i - \bar{x}_{t+1}^\ell \right)^2
\end{align*}
where $x_{t+1}^\ell$ is a sequence of values $x_{t+1}, \ldots,
x_{t+\ell}$, and $\bar{x}_{t+1}^\ell$ is the arithmetic mean of
those values.

The weighted probability of a permutation is defined as:
\begin{align*}
  P_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1}^\ell) \cdot \delta(\phi(x_{t+1}^\ell), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1}^\ell)}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively,
this weighted probability enhances permutations that are involved in
``large'' features and de-emphasizes permutations that are small in
amplitude relative to the features of the time series. Using the
standard form of an entropy, the weighted permutation entropy is then:
\begin{align*}
  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P_w(\pi) \log_2 P_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(\ell!)$, making $0
\le \textrm{WPE} \le 1$.  This normalization is used in all the
results that follow.

In practice, calculating permutation entropy and weighted permutation
entropy involves choosing a good value for the word length $\ell$. The
primary consideration here is that the value be large enough that
forbidden ordinals are discovered, yet small enough that reasonable
statistics over the ordinals are
gathered.  If an average of 100 counts per ordinal is considered to be
sufficient, for instance, then $\ell =
\operatornamewithlimits{argmax}_n \{ T \gtrapprox 100 n! \}$.  In the
literature, $3 \le \ell \le 6$ is a standard choice---generally
without any formal justification.  In theory, the permutation entropy
should reach an asymptote with increasing $\ell$, but that can require
an arbitrarily long time series. In practice, what one should do is
calculate the \emph{persistent} permutation entropy by increasing
$\ell$ until the result converges, but data length issues can intrude
before that convergence is reached.  We used this approach to choose
$\ell = 6$ for the experiments in the following section.  This value
represents a good balance between accurate ordinal statistics and
finite-data effects.

>>>>

Utilizing Shannon entropy rate as a measure of temporal complexity is
not a new idea \cite{Shannon1951, mantegna1994linguistic}, but
previous approaches required categorical data: $x_i \in \mathcal{S}$
for some finite or countably infinite \emph{alphabet} $\mathcal{S}$.
Data taken from real-world systems is, however,
effectively\footnote{Measurements from finite-precision sensors are
  discrete, but data from modern high-resolution sensors are, for the
  purposes of entropy calculations, effectively continuous.}
real-valued.  To analyze real-valued data using a method that requires
categorical values, one must discretize the data---typically by
binning.  Unfortunately, this is rarely a good solution, as the
binning of the values introduces spurious dynamics~\cite{bollt2001}.
The field of symbolic dynamics offers discretization methods that do
not disturb the intrinsic behavior.  These methods are, however,
fragile in the face of noise; worse yet, they require knowledge of the
underlying system.  This is not useful in our context, where we want
to measure the entropy to quantify predictability.

@@

  Given a time series $\{x_i\}_{i = 1,\dots,N}$. Define $\mathcal{S}_\ell$ as all $\ell!$ permutations $\pi$ of order $\ell$. For each $\pi \in \mathcal{S}_\ell$ we determine the relative frequency of that permutation occurring in $\{x_i\}_{i = 1,\dots,N}$:
  \begin{align*}
    P(\pi) = \frac{\left|\{i|i \leq N-\ell,\phi(x_{i+1},\dots,x_{i+\ell}) = \pi\}\right|}{N-\ell+1}
  \end{align*}
  where $P(\pi)$ quantifies the probability of an ordinal and
  $|\cdot|$ is set cardinality. The \emph{permutation entropy} of
  order $\ell \ge 2$ is defined as
  \begin{align*}
    H(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P(\pi) \log_2 P(\pi)
  \end{align*}

@@

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  The intent behind the weighting
is to correct for observational noise that is larger than the trends
in the data, but smaller than the larger scale features.  Consider,
for example, a signal that switches between two fixed points and
contains some additive noise. The PE is dominated by the noise about 
the fixed points driving it to be $\approx 1$, in some sense hiding
the fact that the signal is actually quite structed. The weighted 
permutation entropy on the other hand would be dominated by the 
switching rather than by the stochastic fluctuations. To accomplish 
this, the \emph{weight} of a permutation is taken into account:
\begin{align*}
  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i}^{i+\ell}
                      \left( x_j - \bar{x}_{i+1}^\ell \right)^2
\end{align*}
where $x_{i+1}^\ell$ is a sequence of values $x_{i+1}, \ldots,
x_{i+\ell}$, and $\bar{x}_{i+1}^\ell$ is the arithmetic mean of
those values.

The weighted probability of a permutation is defined as:
\begin{align*}
  P_w(\pi) = \frac{\displaystyle \sum_{i \le N - \ell} w(x_{i+1}^\ell) \cdot \delta(\phi(x_{i+1}^\ell), \pi) }{\displaystyle \sum_{i \le N - \ell} w(x_{i+1}^\ell)}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively,
this weighted probability enhances permutations that are involved in
``large'' features and de-emphasizes permutations that are small in
amplitude relative to the features of the time series. Using the
standard form of an entropy, the weighted permutation entropy is then:
\begin{align*}
  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P_w(\pi) \log_2 P_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(\ell!)$, making $0
\le \textrm{WPE} \le 1$.  This normalization is used in all the
results that follow.

In practice, calculating permutation entropy and weighted permutation
entropy involves choosing a good value for the word length $\ell$. The
primary consideration here is that the value be large enough that
forbidden ordinals are discovered, yet small enough that reasonable
statistics over the ordinals are
gathered.  If an average of 100 counts per ordinal is considered to be
sufficient, for instance, then $\ell =
\operatornamewithlimits{argmax}_{\hat{\ell}} \{ N \gtrapprox 100 \hat{\ell}! \}$.  In the
literature, $3 \le \ell \le 6$ is a standard choice---generally
without any formal justification.  In theory, the permutation entropy
should reach an asymptote with increasing $\ell$, but that can require
an arbitrarily long time series. In practice, what one should do is
calculate the \emph{persistent} permutation entropy by increasing
$\ell$ until the result converges, but data length issues can intrude
before that convergence is reached.  We used this approach to choose
$\ell = 6$ for the experiments in the following section.  This value
represents a good balance between accurate ordinal statistics and
finite-data effects.
